When does vapor pressure deficit drive or reduce evapotranspiration?

### 2017-10-04 doc/paper/vpd_et_paper.tex
think I have final plot 7 ready
okay heading home to cook. first draft done now just need to shell check, proofread, format nicely. then I can send to Pierre

%% To submit your paper:
\documentclass[draft,linenumbers]{agujournal}
\draftfalse

@@

Here we use a Penman-Monteith framework to quantify plant reponse to perturbations to atmospheric demand for water. Section 2 derives the framework, Section 3 describes the data used, Section 4 presents results, and Section 5 discusses conclusions.

@@

We use data from FLUXNET2015. Because $g_1$ coefficients \citep{Lin_2015} and uWUE were only both available for five plant functional types (PFTs - see Table 2),  only 56 of the 77 sites were used. Figure \ref{map_fig} presents each site and its plant functional type.

@@

\section{Results}

@@

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig03.png}
\caption{The joint distribution of $D$ and LAI. LAI has only a weak dependence on $D$}
\label{lai_vpd_fig}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig05.pdf}
\caption{}
\label{term3}
\end{figure}

@@

Table 3 confirms these expectations for PFT behaviour of $\frac{\partial \; ET}{\partial \; D}$. For all PFTs except for CRO, average $\frac{\partial \; ET}{\partial \; D}$ is less than zero. However, $\frac{\partial \; ET}{\partial \; D}$ evaluated at the average of all variables (e.g. LAI, $T$, $c_s$, $D$) is only negative for CSH and GRA. And, DBF in addition to CRO experiences $\frac{\partial \; ET}{\partial \; D}$ < 0 less than half the time. These observations highlight the effect of the nonlinear function in Figure \ref{term3}: $\frac{\partial \; ET}{\partial \; D}$ has a much steeper slope when the function is negative, and is thus more likely to be large magnitude.

The units of $\frac{\partial \; ET}{\partial \; D}$ make it difficult to interpret if $D$ is even a meaningful contributor to ET's variability. To understand $D$'s contribution better, we use a linear approximation and present $\frac{\partial \; ET}{\partial \; D}$ multiplied by $D$'s standard deviation. The range of $D$'s contribution to ET's variability ranges between 16 - 20 W m$^{-2}$ for all PFTs except for CSH, which is about 51 W m$^{-2}$. Another meaningful comparison is to $\frac{\partial \; ET}{\partial \; R} * std(R)$, as net radiation generally the driver of ET (cite joe berry here). For all PFTs except for CSH $D$ contributes between 14.5 - 20.5 \% of $R$'s contribution to variability. For CSH the portion is much larger, about 44 \%. However it is important to note that a linear approximation about a mean base state is probably not a very good approximation across the range of variabilility, so these values are just estimates of $D$'s contribution to ET's variability.

@@

\Begin{table}
\caption{Statistics of $\frac{\partial \; ET}{\partial \; D}$ as a function of PFT.}
\centering
\begin{tabular}{l c c c c c}
  \hline
PFT & $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ & $\frac{\partial \; ET}{\partial \; D}\left(\overline{T, \ldots , D}\right)$ & $\frac{\partial \; ET}{\partial \; D}\left(\overline{T, \ldots , D}\right)*\text{std}(D)$ & $\frac{\frac{\partial \; ET}{\partial \; D}\left(\overline{T, \ldots , D}\right)*\text{std}(D)}{ \frac{\partial \; ET}{\partial \; R}\left(\overline{T, \ldots , D}\right)*\text{std}(R)}$ & fraction $\frac{\partial \; ET}{\partial \; VPD} < 0.$ \\
  \hline
CRO & 0.000853 & 0.026241 & 18.523659 & 0.203022 & 0.473311\\
CSH & -0.108234 & -0.091526 & 50.861613 & 0.439379 & 0.931660\\
DBF & -0.012727 & 0.013794 & 19.734435 & 0.164241 & 0.461674\\
ENF & -0.034087 & 0.000706 & 16.611852 & 0.148548 & 0.534425\\
GRA & -0.019637 & -0.000921 & 16.798083 & 0.173552 & 0.631735\\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}

@@

\subsection{Direct observations of $\frac{\partial \; ET}{\partial \; D}$}

Now that we have an intuitive understanding of $\frac{\partial \; ET}{\partial \; D}$'s behavior, we are equipped to intepret fully realistic plots of $\frac{\partial \; ET}{\partial \; D}$ for each PFT. Figure \ref{fig} presents calculated $\frac{\partial \; ET}{\partial \; D}$ where, unless otherwise noted, all variables in Equation \ref{d_et} are allowed to vary.

%%
 \begin{linenomath*}
 \begin{equation}
 \end{equation}
 \end{linenomath*}

%% Enter Figures and Tables near as possible to where they are first mentioned:
%
% DO NOT USE \psfrag or \subfigure commands.
%
% Figure captions go below the figure.
% Table titles go above tables;  other caption information
%  should be placed in last line of the table, using
% \multicolumn2l{$^a$ This is a table note.}
%
%----------------
% EXAMPLE FIGURE
%
% \begin{figure}[h]
% \centering
% when using pdflatex, use pdf file:
% \includegraphics[width=20pc]{figsamp.pdf}
%
% when using dvips, use .eps file:
% \includegraphics[width=20pc]{figsamp.eps}
%
% \caption{Short caption}
% \label{figone}
%  \end{figure}
%
% ---------------
% EXAMPLE TABLE
%
% \begin{table}
% \caption{Time of the Transition Between Phase 1 and Phase 2$^{a}$}
% \centering
% \begin{tabular}{l c}
% \hline
%  Run  & Time (min)  \\
% \hline
%   $l1$  & 260   \\
%   $l2$  & 300   \\
%   $l3$  & 340   \\
%   $h1$  & 270   \\
%   $h2$  & 250   \\
%   $h3$  & 380   \\
%   $r1$  & 370   \\
%   $r2$  & 390   \\
% \hline
% \multicolumn{2}{l}{$^{a}$Footnote text here.}
% \end{tabular}
% \end{table}

>>>>

%% To submit your paper:
\documentclass[draft,linenumbers]{agujournal}
% \draftfalse
\drafttrue

@@

Here we use a Penman-Monteith framework to quantify plant reponse to perturbations to atmospheric demand for water. Section 2 derives the framework, Section 3 describes the data used, Section 4 presents results, and Section 5 discusses conclusions. The goal of this paper is to use reasonable approximations as a tool to increase intuition for plant response to atmospheric drying. This intuition will aid interpretation of observations and full complexity climate models.

@@

We use data from FLUXNET2015. Because $g_1$ coefficients \citep{Lin_2015} and uWUE were only both available for five plant functional types (PFTs - see Table 2),  only 56 of the 77 sites were used. Figure \ref{map_fig} \explain{map needs to be fleshed out} presents each site and its plant functional type.

@@

\section{Results}
\label{results}

@@

% below show histogram of site specific spearmannr? think would be useful 
\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig03.png}
\caption{The joint distribution of $D$ and LAI. LAI has only a weak dependence on $D$}
\label{lai_vpd_fig}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig05.pdf}
\caption{Sources of variability for Term 2 - Term 3. Top: Term 2 - Term 3 as a function of VPD, with LAI held fixed at PFT averages. The observed range of VPD for each PFT is also shown below the x-axis. Line extent correspondes to 5th and 95th precentiles, while stars denote the location of the 25th, 50th, and 75th percentiles.

Bottom: The location of the minima of ET, as a function of VPD and LAI. Lines and stars denote the distribution of VPD and LAI next to axis, following the same percentiles as above.}
\label{term3}
\end{figure}

@@

Table 3 confirms these expectations for PFT behaviour of $\frac{\partial \; ET}{\partial \; D}$. For all PFTs except for CRO, average $\frac{\partial \; ET}{\partial \; D}$ is less than zero. However, $\frac{\partial \; ET}{\partial \; D}$ evaluated at the average of all variables (e.g. LAI, $T$, $c_s$, $D$) is only negative for CSH and GRA. And, DBF in addition to CRO experiences $\frac{\partial \; ET}{\partial \; D}$ < 0 less than half the time. These observations highlight the effect of the nonlinear function in Figure \ref{term3}: $\frac{\partial \; ET}{\partial \; D}$ has a much steeper slope when the function is negative, and is thus more likely to be large magnitude.
n
The units of $\frac{\partial \; ET}{\partial \; D}$ make it difficult to interpret if $D$ is even a meaningful contributor to ET's variability. To understand $D$'s contribution better, we use a linear approximation and present $\frac{\partial \; ET}{\partial \; D}$ multiplied by $D$'s standard deviation. The range of $D$'s contribution to ET's variability ranges between 16 - 20 W m$^{-2}$ for all PFTs except for CSH, which is about 51 W m$^{-2}$. Another meaningful comparison is to $\frac{\partial \; ET}{\partial \; R} * std(R)$, as net radiation generally the driver of ET (cite joe berry here). For all PFTs except for CSH $D$ contributes between 14.5 - 20.5 \% of $R$'s contribution to variability. For CSH the portion is much larger, about 44 \%. However it is important to note that a linear approximation about a mean base state is probably not a very good approximation across the range of variabilility, so these values are just estimates of $D$'s contribution to ET's variability.

@@

\begin{table}
\caption{Statistics of $\frac{\partial \; ET}{\partial \; D}$ as a function of PFT.}
\centering
\begin{tabular}{l c c c c c}
  \hline
PFT & $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ & $\frac{\partial \; ET}{\partial \; D}\left(\overline{T, \ldots , D}\right)$ & $\frac{\partial \; ET}{\partial \; D}\left(\overline{T, \ldots , D}\right)*\text{std}(D)$ & $\frac{\frac{\partial \; ET}{\partial \; D}\left(\overline{T, \ldots , D}\right)*\text{std}(D)}{ \frac{\partial \; ET}{\partial \; R}\left(\overline{T, \ldots , D}\right)*\text{std}(R)}$ & fraction $\frac{\partial \; ET}{\partial \; VPD} < 0.$ \\
  \hline
CRO & 0.000853 & 0.026241 & 18.523659 & 0.203022 & 0.473311\\
CSH & -0.108234 & -0.091526 & 50.861613 & 0.439379 & 0.931660\\
DBF & -0.012727 & 0.013794 & 19.734435 & 0.164241 & 0.461674\\
ENF & -0.034087 & 0.000706 & 16.611852 & 0.148548 & 0.534425\\
GRA & -0.019637 & -0.000921 & 16.798083 & 0.173552 & 0.631735\\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}

@@

\subsection{Full observations of $\frac{\partial \; ET}{\partial \; D}$}

Now that we have an intuitive understanding of $\frac{\partial \; ET}{\partial \; D}$'s behavior, we are equipped to intepret fully realistic plots of $\frac{\partial \; ET}{\partial \; D}$ for each PFT. Figure \ref{real} presents calculated $\frac{\partial \; ET}{\partial \; D}$ where, unless otherwise noted, all variables in Equation \ref{d_et} are allowed to vary. Each column is a different quantity related to $\frac{\partial \; ET}{\partial \; D}$, and each row is a different PFT.

The full observations generally confirm expectations from Section \ref{results}. CRO has the most postitive values of $\frac{\partial \; ET}{\partial \; D}$, $\frac{\partial \; ET}{\partial \; D}$ is almost always negative for CSH, and reponse depends more with the environmental conditions for the other PFTs (especially ENF). Through the columns of Figure \ref{real} we can see the impact of LAI and $g_a$ on the variability of $\frac{\partial \; ET}{\partial \; D}$. $g_a$'s (columns 1 and 3) scaling alters the magnitude considerably. $LAI$ (columns 2 and 3) variability adds a lot of additional noise to the signal of $\frac{\partial \; ET}{\partial \; D}$, which is slightly undesirable given LAI's role in representing model and observational uncertainty. However, the general story with the noise appears to match the cleaner signal when $LAI$ is help constant and $D_{ETmin}$ is clearly visible \explain{I really need to make these pltos better - way too much overlapping of points that hurts the story}. One exception is possibly with GRA, for which uncertainty is high and causes the full complexity plots (Columns 1 and 2) to not match well with LAI held fixed (Columns 3 and 4).

For ENF and GRA $D_{ETmin}$ does not appear to be only a function of LAI. It turns out that the site to site variaibility in $\gamma$ causes $D_{ETmin}$ to vary, which is not discussed in the pregious section. The impact is observable in both ENF and GRA, but especially for ENF which has a larger $\frac{\partial^2 \; ET}{\partial^2 \; D}$ than the other PFTs.

In general the full complexity plots of $\frac{\partial \; ET}{\partial \; D}$ match our expectations, even with the large sensitivty to LAI measures of uncertainty observed in Figure \ref{term3}. Our LAI-based method of uncertainty propogation blurrs the idealized expectations the most for GRA, and also has a considerable effect for CRO. We therefor have the most confidence in our conclusion based on Equation \ref{d_et} for PFTS CSH, DBF, and ENF, as the full complexity plots with uncertainty included closely match the story when LAI is held fixed. **see somewhat preferred alternate figure \ref{real2} \explain{I think I like this more as thinking in terms of T and RH is easier. However, I used the other plot because Fig 5 does not discuss things in terms of temperature, as this would make things more complicated (adding another diminsion). I could just include both versions of figure 6 though (using figure 6a as a bridge to figure 6b)}.
\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{./fig06.png}
\caption{Scatter plots of $\frac{\partial \; ET}{\partial \; D}$. Each row is a different PFT, and each column is a different quantity related to $\frac{\partial \; ET}{\partial \; D}$, as labelled: Column 1 - $\frac{\partial \; ET}{\partial \; D}$; Column 2 - $\frac{\partial \; ET}{\partial \; D}$ normalized by $g_a$; Column 3 - $\frac{\partial \; ET}{\partial \; D}$ with LAI held fixed at PFT average; and Column 4 - $\frac{\partial \; ET}{\partial \; D}$ normalized by $g_a$ and with LAI held fixed. For reference, lines corresponding to RH = 20\% and RH = 90 \% are drawn. Please note differences in the colorbar scale.}
\label{real}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{./fig06b.png}
\caption{****alternate Fig 06****  Scatter plots of $\frac{\partial \; ET}{\partial \; D}$. Each row is a different PFT, and each column is a different quantity related to $\frac{\partial \; ET}{\partial \; D}$, as labelled. If I end up using this, I could also draw on the curve of $D_{ETmin}$ with $\overline{LAI}$. }
\label{real2}
\end{figure}

\section{Conclusions} \explain{also need to flesh this section out}

The idealized representation of ET used here is successful in developing intuition for how ET responds to changes in $D$. This intuition will aid the community in interpreting observations and output from sophisticated full complexity climate models.

The idealized framework leads to the following general conclusions:
\begin{itemize}
  \item Aerodynamic resistance plays an important role of scaling $\frac{\partial \; ET}{\partial \; D}$. This is a leading order effect for observing higher magnitude reponses in DBF and ENF.
\item In general, CSH has the most negative (i.e. ET reduced) response to increases in $D$ (atmospheric drying). So CSH plants will almost always try and conserve water, effectivey reducing ET with dry atmospheric perturbation.
\item CRO has the most postive response (i.e. ET increased) in response to increases in $D$. This is consistent with CROs that may be evolved or bred to thrive in non-water-limited environments.
\item The reponse is more a function of the environment for DBF, ENF, and GRA. Because as VPD increases the response is more likely to be postive, if RH is fixed then the reponse will be more likely to be positive at warmer T, or if T is fixed the response is more likely to be positive with decreasing RH.
\item ENF has the strongest dependence on environmental conditions due to its small $g1$.
\item Model and observational uncertainty is highest for GRA and CRO, so conlusions about those PFTs should be tempered.
\item However, inclusion of uncertainty doesn't alter conclusions about DBF, ENF, and CSH.
\end{itemize}

The intuition developed using this framework can be used to understand how the land surface will respond and contribute to changes in the environment.

%%
%% Enter Figures and Tables near as possible to where they are first mentioned:
%
% DO NOT USE \psfrag or \subfigure commands.
%
% Figure captions go below the figure.
% Table titles go above tables;  other caption information
%  should be placed in last line of the table, using
% \multicolumn2l{$^a$ This is a table note.}
%
%----------------
% EXAMPLE FIGURE
%
% \begin{figure}[h]
% \centering
% when using pdflatex, use pdf file:
% \includegraphics[width=20pc]{figsamp.pdf}
%
% when using dvips, use .eps file:
% \includegraphics[width=20pc]{figsamp.eps}
%
% \caption{Short caption}
% \label{figone}
%  \end{figure}
%
% ---------------
% EXAMPLE TABLE
%
% \begin{table}
% \caption{Time of the Transition Between Phase 1 and Phase 2$^{a}$}
% \centering
% \begin{tabular}{l c}
% \hline
%  Run  & Time (min)  \\
% \hline
%   $l1$  & 260   \\
%   $l2$  & 300   \\
%   $l3$  & 340   \\
%   $h1$  & 270   \\
%   $h2$  & 250   \\
%   $h3$  & 380   \\
%   $r1$  & 370   \\
%   $r2$  & 390   \\
% \hline
% \multicolumn{2}{l}{$^{a}$Footnote text here.}
% \end{tabular}
% \end{table}
